---
title: "Getting started with datareceipt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with datareceipt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# These chunks talk to the real site, so they only run where a key is set.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("DATARECEIPT_API_KEY"))
)
```

[Data Receipt](https://datareceipt.io) is for asking people for data in a shape you choose. You define a request (a set of columns, each with a type and rules), share its link, and every spreadsheet uploaded against it is validated before it is accepted. This package is the other end of that: it brings the accepted data into R.

## Your key

Every call is authenticated with an API token that belongs to your account. Create one on the site under **Settings > API tokens**, then save it once with `install = TRUE`:

```{r, eval = FALSE}
library(datareceipt)

datareceipt_api_key("12|dr_...", install = TRUE)
```

This appends `DATARECEIPT_API_KEY` to your `.Renviron`, so R finds it in every future session. Restart R (or run `readRenviron("~/.Renviron")`) and check:

```{r}
library(datareceipt)

datareceipt_whoami()
```

## What have I asked for?

`list_requests()` is a tibble of your requests, newest first, with how many submissions and rows each has received:

```{r}
requests <- list_requests()
requests
```

Each request's spec is in the `columns` list column, or ask for it directly. This is the shape your data will come back in, so it is worth a look before you start:

```{r}
get_columns(requests$id[[1]])
```

`type` is one of `text`, `numeric`, `integer`, `date`, or `boolean`; `required` means no empty cells were allowed; `min`, `max`, and `allowed_values` are the rules senders had to meet. `on_break` says what happens to a value that breaks them: `block` stops the submission, `flag` accepts the value and marks it (more on that below). `friendly_name` and `description` are the plain-language name and note the request's owner wrote for the column, as shown in the data dictionary.

## The data

`get_data()` returns every accepted row across every submission to a request, as one tibble. Pass the id, or paste the request's URL from your browser:

```{r}
data <- get_data(requests$id[[1]])
data
```

Columns are typed from the spec: text is character, numbers are double, whole numbers are integer, dates are `Date`, yes/no is logical, and an empty cell is `NA`. The first columns say where each row came from: `submission_id`, `row` (its position within that submission), `submitted_at`, `sender_name`, and `sender_email`. Leave the sender columns out with `include_sender = FALSE`.

From here it is ordinary tidyverse work:

```{r, eval = FALSE}
library(dplyr)

data |>
  summarise(rows = n(), .by = sender_name)
```

## Flagged cells

A column set to `flag` accepts a value that breaks its rules and keeps it as the sender typed it. In `get_data()` those cells are cast where the text allows it, so an out-of-range `"200"` in a whole-number column is still `200L`, and are `NA` where it does not, like `"abc"` in the same column. `get_data()` tells you how many cells were flagged, and `get_flags()` lists them: where each one is, the text as sent, and the rule it broke.

```{r}
get_flags(requests$id[[1]])
```

## One submission at a time

`list_submissions()` shows who sent what, how (`source`: an uploaded `xlsx` or `csv`; submissions from August 2026 may read `paste`, `typed`, or `form`, from sender methods the site no longer offers), and when, along with how many cells were flagged and whether the request's owner has since edited the submission on the site (`revised_at`). `get_submission()` fetches one submission's rows in the same shape as `get_data()`:

```{r}
submissions <- list_submissions(requests$id[[1]])
submissions

get_submission(submissions$id[[1]])
```

## When something goes wrong

Errors carry the site's message and a hint. A 401 means the key is missing, wrong, expired, or revoked; a 404 means nothing with that id exists in the account the key belongs to. Set `options(datareceipt.verbose = TRUE)` to print each request URL as it is made.
